---
title: "Phylogeny vs. popularity in the birds of New Zealand"
author: "Amelia Barber"
date: "11/24/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What is the relationship between bird phylogeny and popularity?

Full blogpost at https://barber.science/2019/11/24/nz-bird-phylogeny-popularity/

```{r}
library(tidyverse)
library(wikitaxa)
library(treeio)
library(ggtree)
library(patchwork)
```

```{r}
nz_bird <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-19/nz_bird.csv")
```

## Extract the common names of the top 30 birds

```{r}
top_birds <- nz_bird %>%
    filter(!is.na(vote_rank) & !is.na(bird_breed)) %>%
    count(bird_breed) %>%
    arrange(desc(n)) %>%
    top_n(30) %>% 
    pull(bird_breed)
```

## Using `wikitaxa` to obtain scientific names from a bird's common name

```{r, cache=TRUE}
bird_list <- top_birds %>% 
    set_names() %>% 
    map(wikitaxa::wt_wikispecies_search)
```

## Pluck hits from Wikispecies results list

```{r}
query_results <- map(bird_list, ~pluck(.x, "query", "search"))
```

## Condense multiple Wikispecies hits down to a single (correct) scientific name

```{r}
query_results_fulldf <- query_results %>% 
    map_dfr(compact, .id = "common_name")

bird_names <- query_results_fulldf %>% 
    filter(!title %in% c("Black robin associated gemykibivirus 1",
                         "Steven Chew Kea Foo",
                         "Thyrocopa kea",
                         "Domene tui",
                         "Avian avulavirus 10",
                         "Haplomitrium ovalifolium",
                         "Cerylinae",
                         "Sibine stimulea",
                         "Homaloptera parclitella",
                         "Mohoua")) %>% 
    distinct(common_name, .keep_all = TRUE) %>%
    select(common_name, latin_name = title) %>%
    mutate(latin_name = str_replace_all(latin_name, 
                                        c("Gemykibivirus" ="Petroica traversi",
                                          "Eudyptula" = "Eudyptula minor",
                                          "Acanthisitta chloris citrina" = "Acanthisitta chloris",
                                          "Ixobrychus" = "Ixobrychus sturmii"
                                          )))
```

```{r}
missing_birds <- keep(query_results, is.null) %>% 
    names()
```

```{r}
missing_bird_names <- tibble(
    common_name = missing_birds,
    latin_name = c("Charadrius bicinctus", "Apteryx australis", "Eudyptes pachyrhynchus")
)
```

```{r}
bird_names_complete <- bind_rows(bird_names, missing_bird_names)
```

## Using NCBI Taxonomy to infer a phylogeny among the most popular birds of New Zealand

Exported my list of newly-assembled Latin names to a text file, and uploaded that into the [NCBI Common Tree](https://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi) web portal

```{r}
plot_set <- bird_names_complete %>% 
    pull(latin_name)

write_lines(plot_set, here("extra_data", "sci_names.txt"))
```

## Using `treeio` to import and manipulate tree data

```{r}
raw_tree <- treeio::read.tree(here::here("extra_data", "phyliptree_fix.phy")) %>% 
    as_tibble() %>%
    mutate(label = str_replace_all(label, "_", " ")) %>% 
    left_join(bird_names_complete, by = c("label" = "latin_name")) %>% 
    mutate(label = if_else(!is.na(common_name), glue::glue("{common_name} ({label})"),
                           label))
tree <- treeio::as.phylo(raw_tree)
```

## Examining taxonomic relationship vs. bird popularity

```{r}
vote_df <- nz_bird %>%
    filter(!is.na(vote_rank) & !is.na(bird_breed)) %>%
    count(bird_breed) %>%
    arrange(desc(n)) %>%
    top_n(30) %>% 
    left_join(bird_names_complete, by = c("bird_breed" = "common_name")) %>% 
    mutate(label = glue::glue("{bird_breed} ({latin_name})")) %>% 
    select(label, n)
```

```{r}
highlight <- vote_df %>% 
    head(3) %>% 
    pull(label)

colors <- c("black", "darkcyan")

p1 <- ggtree(tree, branch.length = "none") +
    geom_nodelab(geom = "label", size = 3) +
    geom_tiplab(aes(color = label %in% highlight), show.legend = FALSE) +
    scale_color_manual(values = colors) +
    coord_cartesian(clip = 'off') +
    ylim(0.5, 29.5) +
    theme_tree2() +
    theme(plot.margin=margin(6, 250, 6, 6),
          axis.line.x = element_line(color = "transparent"),
          axis.ticks.x = element_line(color = "transparent"),
          axis.text = element_text(color = "transparent"))

```

## Generate bar plot

```{r}
p2 <- p1 %>% 
    filter(isTip) %>% 
    select(label, y) %>%
    left_join(vote_df) %>%
    mutate(highlight = if_else(n > 8000, "yes", "no")) %>% 
    ggplot(aes(y, n, fill = highlight)) + 
    geom_col(show.legend = FALSE) + 
    scale_fill_manual(values = colors) +
    coord_flip() +
    annotate("text", x = 0.5, y = 12500, label = "Total # Votes", hjust = 1) +
    xlim(0.5, 29.5) +
    theme_tree2()
```

## Combine using patchwork

```{r, fig.width=8, fig.height=7}
p1 + p2 + plot_annotation(
    title = "Are closely related New Zealand birds equally popular?",
    subtitle = glue::glue("Among the 30 most popular birds in the 'NZ Bird of the Year competition'
                          The three bird species receiving the most votes are highlighted in teal."),
    caption = "Data: NZ Forest and Bird Organization\n Visualization @frau_dr_barber",
    theme = cowplot::theme_map()
    )

ggsave(here::here("plots", "nz_birds.png"), width = 10, height = 7)
```

### So, are the most popular birds closely related?

Not really. Looking at the plot, there doesn't seem to be a clear relationship between where a species is in the phylogeny and the number of votes they received. The three birds receiving the most votes belong to different families. The overall winner, the yellow-eyed penguin is in the _Spheniscidae_ family, the second place kākāpō (of [R4DS](https://r4ds.had.co.nz/) cover fame) is in the _Psittacidae_ family, while the third place black robin is a _passeriformes_.

My guess is that bird popularity is mostly a factor of bird cuteness, which is an evolutionarily [convergent trait](https://en.wikipedia.org/wiki/Convergent_evolution), rather than arising from a particilar branch of the tree.